Measurement Bias from Address Aliasing

### 2016-05-15 presentation.tex
Fill in shim of remaining slides

\begin{itemize}
  \item What is measurement bias?
  \item What is address aliasing?
  \item How does this help?
\end{itemize}

@@

What is the execution time of this function?

@@

\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}

\begin{itemize}
  %\item Collisions between static and automatic variables, depending on alignment of stack.
  \item Address of i, j and k are fixed for all environments.
  \item Address of g and inc is pushed down with length of environment variables.
  \item Spike when inc is at 0x7fffffffe03c + $ k \times $ 0x1000, \emph{aliasing} with i in the last 12 bits.
  \item Performance impact from \emph{false dependencies} in CPU out of order execution.
  \item Effect known as \emph{address aliasing}.
\end{itemize}

\end{column}
\hfill
\begin{column}{.48\textwidth}

% Virtual memory/randomization illustration
\begin{figure}
  \begin{tikzpicture}[font=\footnotesize]
    % See page 453 in pgf manual
    \node [
      rectangle split, rectangle split parts=5, 
      rectangle split part fill={white, white, lightgray, white, lightgray},
      draw, anchor=center, text width=2.5cm
    ] (m) {
      \nodepart{one}
        \scriptsize{USER=lars;SHELL
        =/bin/bash;PWD=
        /home (...)}
      \nodepart{two}
        .stack
        \begin{flushright}
        inc \\
        g
        \end{flushright}
      \nodepart{four}
        .bss
        \begin{flushright}
        k \\
        j \\
        i
        \end{flushright}
    };

@@

    %\node[right] at (m.north east) { \scriptsize{0x7fff'ffffffff} } ;
    %\node[right] at (m.south east) { \scriptsize{0x400000} } ;
  \end{tikzpicture}
  \caption{4K address aliasing between stack and static data.}
\end{figure}

@@

  \begin{frame}
    \frametitle{Bias in Heap Allocation}

@@

    Looked at several allocators:

    \begin{itemize}
      \item ptmalloc (glibc)
      \item tcmalloc (Google)
      \item jemalloc
      \item Hoard
    \end{itemize}

@@

\begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth]
$ man mmap

@@

@@

@@

% Convolution example code
\begin{figure}[t]
  \lstinputlisting[
    language=C,
    frame=single,
    xleftmargin=.00\textwidth,
    xrightmargin=.00\textwidth
  ]{bin/convolution-kernel.c}
  \caption{Simplified implementation of convolution with a fixed kernel. This program is highly sensitive to aliasing between input and output arrays, continuously generating false dependencies.}
  \label{lst:conv}
\end{figure}

@@

% The most awesome graph, O2 of convolution kernel
\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
\begin{figure}[t]
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O2,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/1.5,
        height=4.7cm,
        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel, input size $n=2^{20}$.}
  %Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
\end{figure}

    % Manual offset code (?)

@@

  \begin{frame}
    \frametitle{Aliasing in BLAS libraries}

@@

>>>>

\begin{itemize}
  \item Identify address aliasing as underlying explanation for measurement bias.
  % \begin{itemize}
  %   \item Looking at example from previous work
  % \end{itemize}
  \item See how aliasing affects dynamic memory allocation.
  \item Identify opportunities to optimize for aliasing.
\end{itemize}

@@

Execute program under different environments, setting dummy variable to $K$ zero characters.

\begin{lstlisting}[breaklines=true, belowskip=-10pt]
$ env -i X=`head -c K </dev/zero | tr '\0' '0'` ./loop
\end{lstlisting}

@@

\begin{columns}[T] % align columns
\begin{column}{.50\textwidth}

\begin{itemize}
  %\item Collisions between static and automatic variables, depending on alignment of stack.
  \item Addresses of static i, j and k are fixed for all environments.
  \item Addresses of g and inc are pushed down with size of environment variables.
  \item Spike when address of inc \emph{aliasing} with i in the last 12 bits.
  \item Performance impact from \emph{false dependencies} in CPU out of order execution.
\end{itemize}

\end{column}
\hfill
\begin{column}{.50\textwidth}

% Virtual memory/randomization illustration
\begin{figure}
  \begin{tikzpicture}[font=\footnotesize]
    % See page 453 in pgf manual
    \node [
      rectangle split, rectangle split parts=5, 
      rectangle split part fill={white, white, lightgray, white, lightgray},
      draw, anchor=center, text width=2.5cm
    ] (m) {
      \nodepart{one}
        \scriptsize{USER=lars;SHELL
        =/bin/bash;X=0000
        00000000000000...}
      \nodepart{two}
        .stack
        \begin{flushright}
        inc \\
        g
        \end{flushright}
      \nodepart{four}
        .bss
        \begin{flushright}
        k \\
        j \\
        i
        \end{flushright}
    };

@@

    %\node[right] at (m.north east) { \scriptsize{0x7fff'ffffffff} } ;
    %\node[right] at (m.south east) { \scriptsize{0x400000} } ;
  \end{tikzpicture}
  \caption{Addresses of each variable in the first spike.}
\end{figure}

@@

\begin{frame}[fragile]
\frametitle{Bias in Heap Allocation}

@@

\begin{table}
  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers. Same 12 bit suffix indicate an aliasing pair.}
  \pgfplotstabletypeset[
    font=\footnotesize,
    col sep=comma,
    string type,
    columns={Allocation, 5120, 1048576},
    column type=r,
    columns/Allocation/.style={
      string type, 
      column type=l,
      column type/.add={|}{}
    },
    every head row/.style={
      output empty row,
      before row={\hline
         & 5,120 B & 1,048,576 B \\
        },
      after row=\hline\hline,
    },
    every last row/.style={after row=\hline},
    every last column/.style={column type/.add={}{|}},
    every odd row/.style={after row=\hline},
  ]{bin/malloc-comparison.csv}
\end{table}

@@

\begin{lstlisting}[frame=single, xleftmargin=.01\textwidth, xrightmargin=.01\textwidth, belowskip=1cm]
$ man mmap

@@

Since page size is usually 4096 byte, such allocations will \emph{always} alias with each other.

@@

Example program which is sensitive to aliasing between buffers, continuously generating false dependencies:

@@

\begin{figure}[t]
% Convolution example code
  \lstinputlisting[
    language=C,
    frame=single,
    xleftmargin=.00\textwidth,
    xrightmargin=.00\textwidth,
    aboveskip=-5pt
  ]{bin/convolution-kernel.c}
  \caption{Simplified implementation of convolution with a fixed kernel.}
  \label{lst:conv}
\end{figure}

@@

% The most awesome graph, O2 of convolution kernel
\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
\begin{figure}[t]
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O2,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/1.5,
        height=4.7cm,
        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel, input size $n=2^{20}$.}
\end{figure}

@@

\begin{frame}[fragile]
\frametitle{Aliasing in BLAS libraries}

@@

Measured matrix-vector multiplication (cblas\_dgemv) with increasing offset between buffer addresses.

\pgfplotstableread{bin/libblas.dat}{\libblastable}
\begin{figure}
  \begin{tikzpicture}
    \begin{axis}[
        title=libblas,
        height=4.7cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/1.5, % Make it fit to text width side by side
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
\end{figure}

\begin{frame}
\frametitle{Optimizing for address alias}

% Effects of address aliasing are real, these are some ideas for optimization:

\begin{itemize}
  \item Mark buffers with restrict
  \item Use a special purpose allocator
  \item Manually adjust address offset
\end{itemize}
